---
title: "Reporting a Bayesian model with apabayes"
shorttitle: "apabayes example"
author:
  - name: A. N. Author
    corresponding: true
    email: author@example.edu
    affiliations:
      - name: Example University
abstract: >
  A minimal apaquarto document showing apabayes in both of its registers:
  a number in a sentence and a table with its note.
floatsintext: true
# This document ships with no fitted models, so every number below comes
# from a stored posterior summary. See the setup chunk.
knitr:
  opts_chunk:
    # knitr's default svg device needs cairo or X11. png needs neither,
    # and the typst format aborts every chunk without it.
    dev: png
format:
  apaquarto-pdf:
    documentmode: man
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
---

```{r}
#| label: setup
#| include: false
library(apabayes)
library(apa7)

# Both objects are stored summaries that ship with apabayes, so this
# document renders without brms, blavaan, or a compiler. The chunk below
# shows what these two lines are in a real manuscript.
posterior <- readRDS(
  system.file("extdata", "parameters.rds", package = "apabayes")
)
fit_indices <- readRDS(
  system.file("extdata", "sem_fit.rds", package = "apabayes")
)

# apa_note() is read by the `apa-note` chunk option below, and a chunk
# option is evaluated before its own chunk runs. So the table has to be
# built *here*, in an earlier chunk, not inside the chunk that prints it.
# This is the first thing that goes wrong when copying this file.
posterior_table <- apa_table(posterior)
fit_table <- apa_table(fit_indices)
```

In your own manuscript the setup chunk fits the model instead, and
everything after it is unchanged:

```{r}
#| eval: false
fit <- brms::brm(mpg ~ wt + am, data = mtcars)
posterior <- apa_tidy(fit)
```

## Numbers in a sentence

`apa_inline()` returns the APA string for one term, formatted for whatever
this document is rendering to. Weight predicted fuel economy,
`r apa_inline(posterior, "wt")`, while the transmission effect was
indistinguishable from zero, `r apa_inline(posterior, "am")`.

Nothing above names a format. `apa_inline()` asks knitr what is being
rendered and spells its symbols accordingly: PDF gets LaTeX maths, because
the PDF font drops a combining circumflex, and Word, HTML and Typst get
Unicode.

## Tables

`apa_table()` takes the same object and returns a data frame that
`apa7::apa_flextable()` renders. `apa_note()` builds the note that defines
whatever the table's columns turned out to be.

```{r}
#| label: tbl-posterior
#| tbl-cap: Posterior Summary for the Regression Model
#| echo: false
#| apa-note: !expr apa_note(posterior_table)
apa7::apa_flextable(posterior_table)
```

The same two calls work on every object apabayes understands. @tbl-fit is a
Bayesian structural equation model's fit indices rather than a regression's
coefficients, and the code is unchanged.

```{r}
#| label: tbl-fit
#| tbl-cap: Fit Indices
#| echo: false
#| apa-note: !expr apa_note(fit_table)
apa7::apa_flextable(fit_table)
```

Model fit was acceptable, `r apa_inline(fit_indices)`.

## What apabayes did not do

Every number in this document came from easystats or from the model's own
package. apabayes selected columns, arranged them and formatted them. It
did not decide whether any of these values is good news.
