---
title: "Reporting brms models in apaquarto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reporting brms models in apaquarto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(apabayes)
```

apabayes turns a fitted Bayesian model into the two things an APA
manuscript needs from it: a number in a sentence, and a table with a note
that says what its columns mean. It does not fit anything, and it does not
compute anything — every value comes from the easystats packages or from
the model's own package. apabayes selects, arranges and formats.

This vignette reports a stored posterior summary rather than fitting a
model, so that it builds in a second and without a compiler. Everything
shown works identically on a live fit.

## One object in the middle

`apa_tidy()` is the only function that knows about model classes. It takes
a `brmsfit`, a `lavaan` or `blavaan` fit, a `brms` hypothesis test, a
`loo` comparison, a Bayes factor object, posterior contrasts, a Bayesian
correlation — fourteen object families in all, listed on `?apa_tidy` —
and returns one shape:

```{r}
#| eval: false
fit <- brms::brm(mpg ~ wt + am, data = mtcars)
posterior <- apa_tidy(fit)
```

```{r}
posterior <- readRDS(
  system.file("extdata", "parameters.rds", package = "apabayes")
)
posterior
```

The class is `apabayes_tidy`, a tibble with the interval method, interval
level and centrality recorded as attributes, so nothing downstream has to
guess whether a bound is a credible interval or a highest-density one.

Everything after this point takes that object, not the model. That is the
whole architecture: one extract layer per model class, and a single
reporting layer that never sees a fit.

## A number in a sentence

`apa_inline()` addresses one term by name and returns the APA string:

```{r}
apa_inline(posterior, "wt")
```

In a Quarto or R Markdown document you write that inline, and the string
lands in the prose:

```
Weight predicted fuel economy, `r knitr::inline_expr('apa_inline(posterior, "wt")')`.
```

The object it returns follows papaja's `apa_results` shape, so
`$estimate`, `$statistic` and `$full_result` are all there if you want a
part rather than the whole.

`ci_label` controls the interval's label, and `NULL` prints the brackets
bare, for a sentence that lists several estimates:

```{r}
apa_inline(posterior, "wt", stats = character(), ci_label = NULL)
```

## The number behind the string

Some sentences need the value rather than the string: a number that goes
into arithmetic, into a comparison, or into wording you format yourself.
`apa_value()` addresses a row exactly as `apa_inline()` does and returns
what is in the column:

```{r}
apa_value(posterior, "wt")
apa_value(posterior, "wt", column = "ci_high")
```

It is the alternative to a hand-rolled lookup helper, which indexes by
position and reports the wrong row the day the table changes. The column
defaults to the value the table's type is about — `estimate` here,
`elpd_diff` on a `loo` table, `bf` on a Bayes-factor one — and nothing is
rounded or marked up, so a value that lands in prose goes through the
format layer on the way:

```{r}
apa_num(apa_value(posterior, "wt") * 2)
```

## A table and its note

`apa_table()` returns a data frame whose columns are already formatted and
whose headers are already APA. `apa_note()` builds the note that defines
them — and it defines whatever the table actually has, not a fixed list:

```{r}
tab <- apa_table(posterior)
names(tab)
apa_note(tab)
```

In an apaquarto document the table is printed by apa7, and the note goes
in the chunk option:

````
```{r}
#| label: tbl-posterior
#| tbl-cap: Posterior Summary
#| apa-note: !expr apa_note(tab)
apa7::apa_flextable(tab)
```
````

One thing to know before you copy that: a chunk option is evaluated
*before* its own chunk runs, so `tab` has to be built in an earlier chunk.
Building it inside the chunk that prints it will fail.

The same two calls work on all nine kinds of table apabayes produces.
Model comparison, for instance:

```{r}
comparison <- readRDS(
  system.file("extdata", "loo.rds", package = "apabayes")
)
apa_table(comparison)
```

`loo::loo_compare()` signs every difference against the model with the
highest ELPD, so a sentence that reports how far a baseline trails would
have to flip the sign by hand. `apa_tidy(x, reference = "baseline")`
takes the differences from that model instead, and the model loo
referenced then prints a positive ΔELPD. The standard errors that
survive are the ones loo measured: a `compare.loo` object does not carry
the pointwise ELPDs any other pair would need, so those are `NA` and the
pair is compared with `loo::loo_compare()` directly.

## Beyond one row per parameter

`apa_table()` targets a model-parameter table: one row per parameter, one
column per statistic. A table shaped by a design factor — one row per
condition, one column per measurement occasion — is not something it
produces on its own, because the tidy contract underneath it is one row
per parameter and nothing in `apa_table()` pivots that. Reshape its
output instead.

`group_rows = TRUE` adds a `Component` column, normally used for random
effects; naming your own design factor in `component` repurposes it as
the column `pivot_wider()` spreads on. Two waves of the same two
parameters, illustrated with made-up numbers:

```{r}
by_wave <- apabayes_tidy(
  data.frame(
    term = c("drift", "drift", "boundary", "boundary"),
    label = c(
      "Drift rate", "Drift rate",
      "Boundary separation", "Boundary separation"
    ),
    component = c("Wave 1", "Wave 2", "Wave 1", "Wave 2"),
    effects = "fixed",
    estimate = c(0.42, 0.51, 1.18, 1.24),
    ci_low = c(0.31, 0.39, 0.98, 1.05),
    ci_high = c(0.53, 0.63, 1.38, 1.43),
    pd = c(1, 1, 1, 1)
  ),
  type = "parameters", centrality = "median",
  ci_method = "eti", ci_level = 0.95
)
wide <- apa_table(by_wave, stats = character(), group_rows = TRUE) |>
  tidyr::pivot_wider(
    id_cols = Predictor, names_from = Component,
    values_from = 3
  )
wide
```

```{r}
#| eval: false
apa7::apa_flextable(wide)
```

`apabayes_tidy()` is the public constructor behind this: it takes a plain
data frame with the contract's column names, so a design table built from
a stored summary — not only from a fresh `apa_tidy()` call — starts from
the same place.

## Four output formats

An APA symbol is not one string. A combining circumflex over an R composes
in Word, HTML and Typst, and the PDF font drops it; LaTeX maths sets it
correctly in PDF and is meaningless everywhere else. So apabayes asks
knitr what is being rendered and spells its symbols to suit:

| apaquarto format | what apabayes emits |
| --- | --- |
| `apaquarto-pdf` | LaTeX maths — `$\hat{R}$`, `$\chi^2$` |
| `apaquarto-docx` | Unicode and pandoc markdown |
| `apaquarto-html` | Unicode and pandoc markdown |
| `apaquarto-typst` | Unicode and pandoc markdown |

You do not have to do anything for this; it is what `apa_inline()` does
when you do not pass `markup`. Pass `markup = "plain"` to get an
unformatted string for a console or a test.

`apaquarto-typst` needs **apaquarto 6.0.0 or newer**, where it compiles
for the first time.

## A document to start from

A complete, minimal apaquarto manuscript ships with the package:

```{r}
#| eval: false
system.file("apaquarto-example", package = "apabayes")
```

It holds `example.qmd` and a README with the two commands that render it.
Copy the directory somewhere of your own, run `quarto add wjschne/apaquarto`
beside it, and render.

## What apa_inline() does not do yet

A real manuscript's prose does more with one estimate than apabayes
currently reaches in one call. Measured against a real manuscript
conversion rather than invented, and scoped for a future release rather
than silently absent:

- **A single statistic without the estimate** — quoting only a Bayes
  factor or a *pd* where the sentence has already named the quantity.
  `apa_inline()` always includes the estimate; take the value with
  `apa_value(t, term, column = "bf10")` and format it with `apa_bf()`
  or `apa_pd()`.
- **`*r*` for every correlation method.** A Bayesian Spearman or Kendall
  correlation prints `*r*`, where APA sets `*r*~s~` or `*ρ*`; the table's
  note names the method the coefficient came from.
- **No route reads a named numeric vector** — the natural shape of
  `blavaan::blavFitIndices()` saved on its own. Build a one-row data
  frame with the contract's column names (`ppp`, `brmsea`,
  `bgammahat`, ...) and pass it to `apabayes_tidy()` instead.

## What apabayes will not do

Nothing apabayes prints judges a model. There is no rule that turns a
Bayes factor into "strong evidence" unless you ask for one, no threshold
on R-hat that produces a verdict, and no fit index that comes with a pass
or a fail. Those are the author's to write, and the author's to defend.
